---
title: "R Notebook"
output: html_notebook
---

Starting with a \#tidytuesday data set from back in 2019 <https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-05-21>

<https://ourworldindata.org/plastic-pollution>

```{r, echo=FALSE}
library(tidyverse)
library(rstanarm)
library(janitor)
```

```{r}
theme_set(theme_light())
```

```{r}
coast_vs_waste <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/coastal-population-vs-mismanaged-plastic.csv") 

mismanaged_vs_gdp <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-mismanaged-plastic-waste-vs-gdp-per-capita.csv") 

waste_vs_gdp <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-plastic-waste-vs-gdp-per-capita.csv") 

coast_vs_waste
mismanaged_vs_gdp
waste_vs_gdp

write_csv(coast_vs_waste, 'data/coast_vs_waste.csv')
write_csv(mismanaged_vs_gdp, 'data/mismanaged_vs_gdp.csv')
write_csv(waste_vs_gdp, 'data/waste_vs_gdp.csv')
```

From: <https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups>

|                                                                                                                                                                                                                                                                                                                                                                                                                                    |
|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| "For operational and analytical purposes, economies are divided among income groups according to 2019 gross national income (GNI) per capita, calculated using the World Bank Atlas method. The groups are: low income, \$1,035 or less; lower middle income, \$1,036 - 4,045; upper middle income, \$4,046 - 12,535; and high income, \$12,536 or more. The effective operational cutoff for IDA eligibility is \$1,185 or less." |

+====================================================================================================================================================================================================================================================================================================================================================================================================================================+ +------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}
income_groups <- c('Low income', 'Lower middle income',
                   'Upper middle income', 'High income')

country_by_inc_group <- readxl::read_xls('./data/CLASS.xls', sheet = 'Groups') %>% 
  clean_names() %>% 
  filter(group_name %in% income_groups) %>% 
  rename(inc_group = group_name,
         inc_group_code = group_code)

country_by_inc_group

```

`waste_vs_gdp` **missing data**:

-   So, in the `waste_vs_gdp` dataset there are some non country entities (i.e. groups of countries such as world bank high income countries) which don't have country codes.

-   There is very little data `plastic_waste_per_cap` (values are missing from more than 99% of rows ). I assume this is due to data only being collected on this relatively recently and the dataset covering the time period from 1800 onwards.

-   Similarly there are many missing values for `gdp_per_cap` (values are missing from more than 70% of rows). Again I'll assume this is a data collection issue.

`country_by_inc_group` **missing data**:

-   all rows are present.

joining the two dataset (`waste_vs_gdp` and `country_by_inc_group` ) **missing data:**

-   there is a small number of countries where an income group can't be found. These look to be small islands that are part of larger countries (e.g. Jersey). So I'm happy that the `country_code` variables in both datasets match up reasonably well.

```{r}
# look at missing data for waste_vs_gdp
waste_vs_gdp %>% 
  visdat::vis_miss()

waste_vs_gdp %>% 
  filter(is.na(country_code)) %>% 
  distinct(country)

# look at missing data for country_by_inc_group
country_by_inc_group %>% 
  visdat::vis_miss()

# look at missing data for waste_vs_gdp joined with country_by_inc_group
waste_vs_gdp_inc <- waste_vs_gdp %>%
  filter(!is.na(country_code)) %>% 
  left_join(country_by_inc_group, by = c("country_code" = "country_code")) %>% 
  select(-country_name)

waste_vs_gdp_inc %>%
  visdat::vis_miss()

waste_vs_gdp_inc %>% 
  distinct(country, country_code, inc_group) %>% 
  filter(is.na(inc_group))
```

So, the only year that data is available on plastic waste per capita is 2010

```{r}
waste_vs_gdp_inc %>% 
  filter(!is.na(plastic_waste_per_cap)) %>% 
  distinct(year)

waste_vs_gdp_inc <- waste_vs_gdp_inc %>% 
  filter(!is.na(plastic_waste_per_cap)) %>% 
  
  mutate(plastic_waste_tot = plastic_waste_per_cap * 
           total_population_gapminder) 

waste_vs_gdp_inc %>% 
  
  ggplot(aes(gdp_per_cap, plastic_waste_per_cap,
             size = total_population_gapminder,
             colour = plastic_waste_tot)) +
  geom_point() +
  scale_x_log10() + scale_y_log10() +
  scale_colour_viridis_c(option = 'magma')
```

```{r}
waste_vs_gdp_inc <- waste_vs_gdp_inc %>% 
  left_join(mismanaged_vs_gdp %>% 
              select(country_code, year,
                     mismanaged_plastic_per_cap)) %>% 
  
  mutate(plastic_waste_tot = plastic_waste_per_cap * 
           total_population_gapminder,
         misman_plastic_waste_tot = mismanaged_plastic_per_cap *
           total_population_gapminder,
         prop_plastic_waste_misman = misman_plastic_waste_tot / 
           plastic_waste_tot) 

waste_vs_gdp_inc %>% 
  
  ggplot(aes(gdp_per_cap, prop_plastic_waste_misman,
             size = plastic_waste_tot,
             alpha = plastic_waste_tot,
             fill = log(total_population_gapminder))) +
  geom_point(shape = 21) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_c(option = 'magma', direction = -1) +
  scale_alpha(range = c(0.5, 0.8))


waste_vs_gdp_inc %>% 
  ggplot(aes(plastic_waste_tot)) +
  geom_density() +
  scale_x_log10()
  
```

```{r}
library(rstanarm)

waste_vs_gdp_inc <- waste_vs_gdp_inc %>% 
  mutate(gdp = gdp_per_cap * total_population_gapminder)

# fit a simple model with one predictor
fit_1 <- stan_glm(log(plastic_waste_tot) ~ log(gdp), 
                  data = waste_vs_gdp_inc,
                  refresh = 0)

# develop the model by adding a second predictor
fit_2 <- stan_glm(
  log(plastic_waste_tot) ~ log(gdp) + log(total_population_gapminder), 
  data = waste_vs_gdp_inc,
  refresh = 0
  )

# unsuprisingly the second model is favoured
loo_1 <- loo(fit_1)
loo_2 <- loo(fit_2)
loo_compare(loo_1, loo_2)

plot(loo_1)
plot(loo_2)
```

```{r}
library(tidybayes)

fit_1 %>% 
  spread_draws(`(Intercept)`)

get_variables(fit_1)

fit_1 %>% 
  tidy_draws()

broom.mixed::tidy(fit_1)

pp_check(fit_1, plotfun = "hist", nreps = 5)

pp_check(fit_1, plotfun = "stat", stat = function(y) max(y))

mean(bayes_R2(fit_2))
mean(loo_R2(fit_2))

loo_R2(fit_2) %>% 
  as_tibble() %>% 
  ggplot(aes(value)) +
  geom_histogram()
```

```{r}
library("bayesplot")
library("rstanarm")
library("ggplot2")

fit <- stan_glm(mpg ~ ., data = mtcars)
posterior <-  
  as_tibble(fit)

plot_title <- ggtitle("Posterior distributions",
                      "with medians and 80% intervals")
mcmc_areas(posterior,
           pars = c("cyl", "drat", "am", "wt"),
           prob = 0.8) + plot_title
```
